Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-05-20 agujournaltemplate.tex
Writing Intro 80%

    % Phytoplankton is important
    Phytoplankton provide the foundation for marine food webs and play a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}.
    Phytoplankton consists of a diverse set of organisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}. The short generation time and high diversity allow phytoplankton communities to adapt quickly to environmental changes, making them highly suitable indicators of the ecosystem state \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
    The structure of the community can be considered an integrator of environmental factors, as it is strongly influenced by temperature, nutrient availability, and light availability \cite{mouw_phytoplankton_2016}. In addition to these bottom-up effects, influences from higher trophic levels, such as zooplankton, also structure the community and affect biodiversity through grazing pressure \cite{banas_adding_2011}. Disentangling the effects driving changes in community composition and biodiversity of phytoplankton is a challenging but crucial task for better understanding our oceans.

    Climate change is predicted to have a lasting impact on phytoplankton communities, shifting the ranges and contributions of functional groups \cite{henson_future_2021}. The drivers and effects of these changes require further investigation, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.

    Coastal margins constitute \qty{\sim 10}{\%} of the total ocean surface area, but are responsible for an estimated \qty{40}{\%} of carbon sequestration to the seafloor \cite{yool_examination_2001, mullerkarger_importance_2005}.

    - Interestingly, the trend for coastal phytoplankton biomass has been far from diminishing.
    Using remote sensing, a global increase in coastal phytoplankton bloom frequency and spatial extent has been observed \cite{dai_coastal_2023}.

    % We need time series in order to understand effects and changes (threats) 
    Time series of biological observations offer valuable insights into how marine ecosystems respond to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}.

    A tropical coastal ecosystem that has been studied intensively is the CARIACO time series, which has involved consistent sampling of physical and biogeochemical parameters in the Cariaco Basin from 1995 to 2017 \cite{muller-karger_scientific_2019}.

    % Describe CARIACO location
    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}. The basin is a continental shelf depression consisting of two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}. The water below the sill is poorly ventilated, which, combined with the high surface productivity and microbial respiration of sinking organic matter, results in permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}. Due to these unique properties, the Cariaco Basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of past climate \cite{hughen1996nature} and serving as a testbed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.

    % Describe general seasonal cycle with functional groups etc.
    A markedly seasonal upwelling cycle is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward. A smaller wind-driven secondary upwelling often occurs in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}. During the upwelling period, the supply of deeper nutrient-rich waters supports high biological productivity, and the phytoplankton community is dominated by diatoms \cite{romero_seasonal_2009}, with dinoflagellates, coccolithophorids, and nanoplankton also contributing. In the low-wind rainy season, the Cariaco Basin exhibits oligotrophic conditions, where the fraction of large phytoplankton cells is diminished, but nano- and picophytoplankton are present in considerable numbers throughout the year \cite{lorenzoni_characterization_2015}.

    % Describe CARIACO changes in ecosystem
    Since the inception of the time series in 1995, the ecosystem of the Cariaco basin has undergone marked changes. \citeA{taylor_ecosystem_2012} analyzed time series data up to 2010 and found significant decadal trends of increase in sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors related to the impact of global warming. The phytoplankton community identified by microscopy showed a drastic reduction in Diatom abundance and a shift toward smaller taxa after 2004, which coincided with a collapse in commercial sardine landings and an increase in mesozooplankton biomass.
    The trends in microscopy community data were partially confirmed and expanded by \citeA{pinckney_phytoplankton_2015}. They observed that the drastic reduction in diatoms counted using microscopy corresponded, but to a lesser degree, to diatom abundance calculated from the pigment composition measured by high-performance liquid chromatography (HPLC). This is most likely explained by a shift to diatom species with smaller cell sizes, which are more difficult to detect and identify using light microscopy \cite{lorenzoni_characterization_2015}. In addition, the pigment data indicate a deepening of the chlorophyll maximum, an increase in diversity, and a reduction in seasonal variability in the community composition. These changes are most likely a response to the reduced upwelling and a deepening of the euphotic zone associated with a decrease in the strength of seasonal trade winds. Interestingly, changes in the phytoplankton community did not result in a significant change in the export of particulate organic carbon (POC) to depth, as measured by sediment traps deployed during the time series \cite{pinckney_phytoplankton_2015}.

    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    %% ADD Duplicate Points here:
    During the initial phase of the time series (1996–2004), the Cariaco Basin underwent a consistent seasonal upwelling cycle, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}. In this period, approximately \qty{\sim 70}{\%} of the annual depth-integrated primary production was observed. During upwelling, the phytoplankton community was dominated by microphytoplankton, primarily consisting of diatoms, dinoflagellates, and coccolithophorids \cite{pinckney_phytoplankton_2015}.

    Following 2004, there was a transition in the physical and biogeochemical regime characterized by a significant reduction in trade winds during the upwelling period, which led to elevated sea surface temperatures and decreased mixing between the surface and deep ocean layers \cite{taylor_ecosystem_2012}.

    Microscopy cell counts indicate the collapse of large phytoplankton, particularly diatoms, which is confirmed to a lesser extent by pigment data analysis \cite{pinckney_phytoplankton_2015}. In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, which are typical of stratified water columns \cite{pinckney_phytoplankton_2015}. During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, coinciding with an increase in zooplankton biomass measured in the Cariaco Basin \cite{taylor_ecosystem_2012}.

    Previous analyses of phytoplankton community data did not include time series data beyond 2011, and \citeA{taylor_ecosystem_2012} rightly concluded that the observed changes cannot "be clearly identified as unidirectional trends driven by anthropogenic climate change or whether they reflect low-frequency natural cycles, such as those driving the AMO."  
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

    % NOw we come in!
    In this study, we document changes in biodiversity and community structure in the Cariaco Basin between 1995 and 2017 and quantify the effects of bottom-up drivers and climate indices on changes in the phytoplankton community. We focus on phytoplankton cell count data obtained through microscopy, as these data have not been fully analyzed in previous studies. We argue that, despite the shortcomings of microscopy methods, they provide a consistent measure of changes in large phytoplankton. Our analysis emphasizes bottom-up drivers of community changes, as these data are available throughout the time series and at a monthly resolution. This study complements the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017. Understanding the regional dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems driven by changes in climate.

@@

AMO has emerged as an important indicator of the climate system and how it affects the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}. The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}. Although regional effecs may vary, a positive phase in the AMO is generally associated with periods of warmer temperatures and a shift of the ITCZ from north to south. This shift of the ITCZ southward increases wind-driven upwelling in the Cariaco basin \cite{taylor_ecosystem_2012}. The threshold responses between functional groups, which relate to a shift in the cell count abundances, show contrasting patterns. Diatoms and cyanobacteria show strong response thresholds for negative AMO anomalies. This is an interesting threshold response shown for the AMO, which, if sufficiently negative, seems to correlate to a reorganization of the physical regime in the Cariaco basin. The abundance of haptophytes shows a strong threshold at positive AMO anomalies, which could be related to the pattern observed from HPLC data of haptophytes mostly exhibiting blooms above the mixed layer and at a lower frequency from other functional groups \cite{pinckney_phytoplankton_2015}.

>>>>

    % Phytoplankton is important
    Phytoplankton provide the foundation for marine food webs and play a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}. The paraphyletic group consists of a diverse set of organisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}. The short generation time and high diversity allow phytoplankton communities to adapt quickly to environmental changes, making them suitable indicators of ecosystem state \cite{alvarez-cobelas_what_1998, barton_anthropogenic_2016, di_cavalho_temporal_2023}.
    Coastal margins constitute \qty{\sim 10}{\%} of the total ocean surface area, but are responsible for an estimated \qty{40}{\%} of carbon sequestration to the seafloor \cite{yool_examination_2001, mullerkarger_importance_2005}. Using remote sensing, a global increase in coastal phytoplankton bloom frequency and spatial extent has been observed in the last 20 years, although regional effects are highly variable \cite{dai_coastal_2023}.
    Climate change is predicted to have a lasting impact on phytoplankton communities, shifting the ranges and contributions of functional groups \cite{boyd_biological_2016, henson_future_2021}. 
    Long-term time series of biological observations are crucial to understanding how marine ecosystems respond to temporary shifts, cyclic processes, and long-term changes in the local physical environment and climate \cite{carstensen_need_2014, henson_observing_2016}. Unfortunately, such time series are rare, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and faces greater threats of extinction compared to temperate regions \cite{finnegan_paleontological_2015}.

    A tropical coastal ecosystem that has been studied intensively is the CARIACO time series, which has involved consistent sampling of physical and biogeochemical parameters in the Cariaco Basin from 1995 to 2017 \cite{muller-karger_scientific_2019}. 
    % Describe CARIACO location
    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}. The basin is a continental shelf depression consisting of two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}. The water below the sill is poorly ventilated, which, combined with the high surface productivity and microbial respiration of sinking organic matter, results in permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}. Due to these unique properties, the Cariaco Basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of past climate \cite{hughen1996nature} and serving as a testbed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.
    % Describe general seasonal cycle with functional groups etc.
    A markedly seasonal upwelling cycle is driven by easterly trade winds, with the highest levels of primary production observed from December through May when the Inter-Tropical Convergence Zone (ITCZ) moves southward. A smaller wind-driven secondary upwelling often occurs in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}. During the upwelling period, the supply of deeper nutrient-rich waters supports high biological productivity, and the phytoplankton community is dominated by diatoms \cite{romero_seasonal_2009}, with dinoflagellates, coccolithophorids, and nanoplankton also contributing. In the low-wind rainy season, the Cariaco Basin exhibits oligotrophic conditions, where the fraction of large phytoplankton cells is diminished, but nano- and picophytoplankton are present in considerable numbers throughout the year \cite{lorenzoni_characterization_2015}.

    % PARAGRAPH THREE - CHANGES IN CARIACO
    Since the inception of the time series in 1995, the ecosystem of the Cariaco Basin has undergone marked changes. \citeA{taylor_ecosystem_2012} analyzed time series data up to 2010 and found significant decadal trends of increasing sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors linked to the impact of global warming. During the initial phase of the time series (1996–2004), the Cariaco Basin underwent a consistent seasonal upwelling cycle, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}. Following 2004, there was a transition in the physical and biogeochemical regime characterized by a significant reduction in trade winds during the upwelling period \cite{taylor_ecosystem_2012}. This strongly affected the ecosystem and phytoplankton community. Microscopy cell counts indicate the collapse of large phytoplankton, particularly diatoms, as confirmed to a lesser extent by pigment data analysis \cite{pinckney_phytoplankton_2015}. In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, which are typical of stratified water columns \cite{lorenzoni_characterization_2015}. An analysis of the pigment data indicates a deepening of the chlorophyll maximum, an increase in diversity, and a reduction in seasonal variability in community composition \cite{pinckney_phytoplankton_2015}. During this phase, the heavily exploited sardine fisheries in coastal Venezuela experienced a collapse, coinciding with an increase in zooplankton biomass measured in the Cariaco Basin \cite{taylor_ecosystem_2012}. Interestingly, changes in the phytoplankton community did not result in a significant change in the export of particulate organic carbon (POC) to depth, as measured by sediment traps deployed during the time series \cite{pinckney_phytoplankton_2015}.

    % PARAGRAPH FOUR - EXPLANATION OF CHANGES // OPEN QUESTIONS!!
    The shift in the community, due to the warming trend and the reduction in upwelling intensities, does not appear to be a permanent regime shift. Data from 2012 until the end of the CARIACO time series in 2017 indicate a return to stronger upwelling intensities and phytoplankton productivity levels similar to those of the late 1990s \cite{muller-karger_scientific_2019}.
    % What drives the changes??? Bottom-up/ Top-down ... CLIMATE
    Previous analyses of the phytoplankton community did not include time series data beyond 2011, and \citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as 
    "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation" (AMO). 
    From this background, we aim to 1) investigate and relate changes in the depth distribution of biomass and community diversity over the entire time series, 2) check the structure and temporal changes of the phytoplankton community and 3) check whether in-situ upwelling metrics or climatic indices are better predicotrs of community change...
    % - microscopy data was not analyzed in detail, particularly towards time series of diversity.
    % Question: Does phytoplankton diversity follow the trends in biomass?
    We aim to investigate and relate changes in the depth distribution of biomass and community diversity over the entire time series.

    % What drives the changes??? Bottom-up/ Top-down ... CLIMATE
    Previous analyses of the phytoplankton community did not include time series data beyond 2011, and \citeA{taylor_ecosystem_2012} cautioned that the observed changes cannot be clearly identified as 
    "unidirectional trends driven by anthropogenic climate change or whether they simply reflect low-frequency natural cycles, such as those driving the Atlantic Multidecadal Oscillation" (AMO).

    The impact of large-scale climatic processes on the observed shift in the community is still unclear. 
    %ENSO MEI v.2
    The El Niño Southern Oscillation (ENSO) has a strong teleconnection to the Caribbean Sea, as during ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}.

    \citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco basin, despite strong teleconnection effects in the tropical Atlantic \cite{enfield_tropical_1997}.

    % AMO??
    AMO has emerged as an important indicator of the climate system and how it affects the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}. The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}.

@@

    Factors act at different scales \cite{wolkovich_temporal_2014}
    We aim to examine correlations and structures in time series data at monthly and annual time scales.

    The structure of the community can be considered an integrator of environmental factors, as it is strongly influenced by temperature, nutrient availability, and light availability \cite{mouw_phytoplankton_2016}. In addition to these bottom-up effects, influences from higher trophic levels, such as zooplankton, also structure the community and affect biodiversity through grazing pressure \cite{banas_adding_2011}.

    - What are the open questions?? Drivers of community change.
    - Zooplankton/Sardines one important aspect, but data coverage is sparse.
    - Climate impacts, local upwelling system, etc.

    - Microscopy data in more detail (use entire depth and time)

    % FINAL PARAGRAPH LEADING TO METHODS
    In this study, we document changes in biodiversity and community structure in the Cariaco Basin between 1995 and 2017 and quantify the relationships between bottom-up drivers, climate indices and the phytoplankton community. We focus on cell count data obtained through microscopy, as these data have not been fully analyzed in previous studies. We argue that, despite the shortcomings of microscopy methods, they provide a consistent measure of changes in large phytoplankton. Our analysis emphasizes bottom-up drivers of community changes, as these data are available throughout the time series and at a monthly resolution. This study complements the work of \citeA{taylor_ecosystem_2012} and \citeA{pinckney_phytoplankton_2015} with a more detailed analysis of biodiversity aspects using all data available from the CARIACO time series up to the final cruise in 2017. Understanding the regional dynamics of the Cariaco Basin is crucial for comprehending the broader ecological changes in tropical marine ecosystems driven by shifts in climate.

    %%%%%%% REASSEMBLE INTRO ABOVE HERE %%%%%%%%%%%%%

    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%AMO has emerged as an important indicator of the climate system and how it affects the ecosystems of the north Atlantic \cite{nye_ecosystem_2014}. The oscillations of the AMO index are strongly correlated with temperature indices in the Caribbean region \cite{stephenson_changes_2014}. % COPIED TO INTRO
Although regional effecs may vary, a positive phase in the AMO is generally associated with periods of warmer temperatures and a shift of the ITCZ from north to south. This shift of the ITCZ southward increases wind-driven upwelling in the Cariaco basin \cite{taylor_ecosystem_2012}. The threshold responses between functional groups, which relate to a shift in the cell count abundances, show contrasting patterns. Diatoms and cyanobacteria show strong response thresholds for negative AMO anomalies. This is an interesting threshold response shown for the AMO, which, if sufficiently negative, seems to correlate to a reorganization of the physical regime in the Cariaco basin. The abundance of haptophytes shows a strong threshold at positive AMO anomalies, which could be related to the pattern observed from HPLC data of haptophytes mostly exhibiting blooms above the mixed layer and at a lower frequency from other functional groups \cite{pinckney_phytoplankton_2015}.
